Agreement Attraction in Turkish: The Case of Genitive Attractors

### 2022-02-10 paper/lcn_revision/manuscript-knitr.tex
lcn-revision

\documentclass[]{interact}

@@

\newcommand{\rev}[1]{{\color{red}#1}}

\begin{document}

@@

However, a potential problem with the stimuli in the \citet{LagoEtAl:2019} study is that the stems of all head nouns such as \textit{e\u{g}itmen-i} (\textit{`instructor'}) in \ref{ex:lago} were consonant-ending, making their possessive forms ambiguous between possessive and accusative case due to case syncretism between them in vowel-ending stems \citep[pp. 66--67]{GokselKerslake:2005}. As a result, the head noun (\textit{'instructor'}) in \ref{ex:ambiguous} can be disambiguated towards possessive case as in \ref{ex:ambiguous_possessive}, or towards accusative case as in \ref{ex:ambiguous_accusative}, where the genitive-marked noun functions as an embedded subject, and the matrix subject is omitted due to pro-drop. 
\rev{We also calculated the relative probability of having an accusative marked noun compared to an possessive marked noun following genitive marking. Data from annotated treebanks from Universal Dependencies v2.9 \citep{TurkEtAl2021,KuzgunEtAl2020,TurkEtAl2019b,SulubacakEtAl2016,Coltekin2015} showed that the relative probability of encountering accusative marking after the genitive-marked noun is \Sexpr{p_acc_given_gen}.}

@@

We recruited \Sexpr{n_subject} undergraduate students to participate in the experiment in exchange for course credit. All participants were native Turkish speakers, with an average age of \Sexpr{mean_age} (range: \Sexpr{min_age} -- \Sexpr{max_age}). The experiment was carried out following the principles of the Declaration of Helsinki and the regulations concerning research ethics at Bo\u{g}azi\c{c}i University. All participants provided informed consent before their participation and their identity \rev{were} completely anonimised.

@@

Prior to the analysis, we removed the data for all participants who failed to show sufficient sensitivity to the effect of grammaticality in singular attractor conditions, i.e., when no agreement attraction was expected. Specifically, we removed all participants for whom the difference in the percentage of \textit{yes} responses between \rev{conditions with singular attractors \ref{ex:expitem-sgsg} and \ref{ex:expitem-sgpl}} fell below the threshold of \Sexpr{accuracy_threshold} percentage points. We also excluded trials in which the participants missed the response deadline. As a result, we excluded \rev{\Sexpr{perc_deletion_exp1}\%} of the trials from our experiment and \Sexpr{perc_deletion_lago}\% of the \citeauthor{LagoEtAl:2019}'s (\citeyear{LagoEtAl:2019}) trials.

@@

In the results section, we provide summaries of the coefficient posterior distributions. For our model, we ran \Sexpr{n_chains} chains with \Sexpr{n_warmup} warm-up iterations and \Sexpr{n_iter-n_warmup} sampling iterations. The data for our study, along with our analysis scripts can be found at \url{https://osf.io/hegmd/}.

@@

Figure \ref{fig:AverageResponses} shows the average proportions of ``acceptable'' responses by experimental condition in our experiment with unambiguous possessive marking, side by side with Lago et al.'s findings. It shows that ungrammatical sentences with plural attractors are rated as acceptable more often (M = \Sexpr{exp1_cur_entry1$M}, SE = \Sexpr{exp1_cur_entry1$SE}) than their counterparts with singular attractors (M = \Sexpr{exp1_cur_entry2$M}, SE=\Sexpr{exp1_cur_entry2$SE}). The magnitude of the effect (\Sexpr{m_effect}) was in line with the findings reported in \citet{LagoEtAl:2019}, where the difference was also 0.11. Accuracy rates for grammatical conditions were nearly equal (M = \Sexpr{exp1_cur_entry3$M} and \Sexpr{exp1_cur_entry4$M}, SE = \Sexpr{exp1_cur_entry3$SE} and \Sexpr{exp1_cur_entry4$SE}, for singular and plural attractors respectively).

<<AverageResponses, message = FALSE, echo=FALSE,  warning = FALSE, dev='cairo_pdf', fig.height=3, fig.align='center', out.width='\\linewidth', fig.pos='hbt!', fig.cap='The average percentage of acceptable responses according to the experimental conditions in our study and \\citet{LagoEtAl:2019}. Error bars signal standard errors calculated following \\citet{Morey:2008,Cousineau:2005}.'>>=

plot(p_avg_resp)

@

Figure \ref{fig:ResponseModel} shows estimates and 95\% credible intervals of a Bayesian GLM with a probit link function. The main effect of grammaticality (\Sexpr{estimate_cGrammatical}) indicates that, on average, participants were quite good at distinguishing between grammatical and ungrammatical sentences. Meanwhile, the negative interaction between grammaticality and attractor number (\Sexpr{estimate_cGrammatical_cAttractorPlural}) indicated a larger difference (positive) effect of plural attractors in ungrammatical conditions, and thus a number agreement attraction effect. There was weak evidence for a negative three-way interaction between the presence of ambiguity, ungrammaticality, and attractor number (\Sexpr{estimate_threeway_interaction}), which was largely driven by differences in the effect of attractor number in \textit{grammatical conditions}, as the magnitude of the effect in the ungrammatical conditions was identical in both experiments (\Sexpr{m_effect_lago}).  
This is consistent with the estimates of the model based on ungrammatical sentences in Figure \ref{fig:ResponseModelUngram}, which show no indication of an interaction between ambiguity and the presence of a plural attractor ((\Sexpr{estimate_ungramResp_cAmbiguity_cAttractorPlural}). It also showed a main effect of plural attractor (\Sexpr{estimate_ungramResp_cAttractorPlural}). Taken together, the coefficients indicated a substantial agreement attraction effect regardless of the presence of local ambiguity.

<<ResponseModel, message = FALSE, echo=FALSE,  warning = FALSE, dev='cairo_pdf', fig.height=3, fig.align='center', out.width='\\linewidth', fig.pos='hbt!', fig.cap='Estimates and 95\\% credible intervals for the probit regression coefficients for the model of responses in our experiment and \\citet{LagoEtAl:2019}.'>>=

suppressWarnings({
p_m_response
})

@

<<ResponseModelUngram, message = FALSE, echo=FALSE, warning = FALSE, dev='cairo_pdf', fig.height=1.5, fig.align='center', out.width='\\linewidth', fig.pos='hbt!', fig.cap='Estimates and 95\\% credible intervals for the probit regression coefficients for the model of responses to \\protect{\\textbf{ungrammatical sentences}} in our experiment and \\citet{LagoEtAl:2019}.'>>=

suppressWarnings({
p_m_ungram_response
})

@

>>>>

\documentclass[]{interact}\usepackage[]{graphicx}\usepackage[]{color}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

\newcommand{\rev}[1]{{\color{red}#1}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

However, a potential problem with the stimuli in the \citet{LagoEtAl:2019} study is that the stems of all head nouns such as \textit{e\u{g}itmen-i} (\textit{`instructor'}) in \ref{ex:lago} were consonant-ending, making their possessive forms ambiguous between possessive and accusative case due to case syncretism between them in vowel-ending stems \citep[pp. 66--67]{GokselKerslake:2005}. As a result, the head noun (\textit{'instructor'}) in \ref{ex:ambiguous} can be disambiguated towards possessive case as in \ref{ex:ambiguous_possessive}, or towards accusative case as in \ref{ex:ambiguous_accusative}, where the genitive-marked noun functions as an embedded subject, and the matrix subject is omitted due to pro-drop. 
\rev{We also calculated the relative probability of having an accusative marked noun compared to an possessive marked noun following genitive marking. Data from annotated treebanks from Universal Dependencies v2.9 \citep{TurkEtAl2021,KuzgunEtAl2020,TurkEtAl2019b,SulubacakEtAl2016,Coltekin2015} showed that the relative probability of encountering accusative marking after the genitive-marked noun is 0.21.}

@@

We recruited 118 undergraduate students to participate in the experiment in exchange for course credit. All participants were native Turkish speakers, with an average age of 20 (range: 18 -- 32). The experiment was carried out following the principles of the Declaration of Helsinki and the regulations concerning research ethics at Bo\u{g}azi\c{c}i University. All participants provided informed consent before their participation and their identity \rev{were} completely anonimised.

@@

Prior to the analysis, we removed the data for all participants who failed to show sufficient sensitivity to the effect of grammaticality in singular attractor conditions, i.e., when no agreement attraction was expected. Specifically, we removed all participants for whom the difference in the percentage of \textit{yes} responses between \rev{conditions with singular attractors \ref{ex:expitem-sgsg} and \ref{ex:expitem-sgpl}} fell below the threshold of 0.1 percentage points. We also excluded trials in which the participants missed the response deadline. As a result, we excluded \rev{3.39\%} of the trials from our experiment and 2.27\% of the \citeauthor{LagoEtAl:2019}'s (\citeyear{LagoEtAl:2019}) trials.

@@

In the results section, we provide summaries of the coefficient posterior distributions. For our model, we ran 4 chains with 1000 warm-up iterations and 1000 sampling iterations. The data for our study, along with our analysis scripts can be found at \url{https://osf.io/hegmd/}.

@@

Figure \ref{fig:AverageResponses} shows the average proportions of ``acceptable'' responses by experimental condition in our experiment with unambiguous possessive marking, side by side with Lago et al.'s findings. It shows that ungrammatical sentences with plural attractors are rated as acceptable more often (M = 0.24, SE = 0.01) than their counterparts with singular attractors (M = 0.14, SE=0.01). The magnitude of the effect (0.10) was in line with the findings reported in \citet{LagoEtAl:2019}, where the difference was also 0.11. Accuracy rates for grammatical conditions were nearly equal (M = 0.92 and 0.91, SE = 0.01 and 0.01, for singular and plural attractors respectively).

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[hbt!]

{\centering \includegraphics[width=\linewidth]{figure/AverageResponses-1}

}

\caption{The average percentage of acceptable responses according to the experimental conditions in our study and \citet{LagoEtAl:2019}. Error bars signal standard errors calculated following \citet{Morey:2008,Cousineau:2005}.}\label{fig:AverageResponses}
\end{figure}

\end{knitrout}

Figure \ref{fig:ResponseModel} shows estimates and 95\% credible intervals of a Bayesian GLM with a probit link function. The main effect of grammaticality ($\hat{\beta}=3.01;$ $CI=[2.76; 3.27];$ $P(\beta<0)< .001$) indicates that, on average, participants were quite good at distinguishing between grammatical and ungrammatical sentences. Meanwhile, the negative interaction between grammaticality and attractor number ($\hat{\beta}=-0.71;$ $CI=[-1.00; -0.44];$ $P(\beta<0)> .999$) indicated a larger difference (positive) effect of plural attractors in ungrammatical conditions, and thus a number agreement attraction effect. There was weak evidence for a negative three-way interaction between the presence of ambiguity, ungrammaticality, and attractor number ($\hat{\beta}=-0.29;$ $CI=[-0.81; 0.20];$ $P(\beta<0)=    .88$), which was largely driven by differences in the effect of attractor number in \textit{grammatical conditions}, as the magnitude of the effect in the ungrammatical conditions was identical in both experiments ().  
This is consistent with the estimates of the model based on ungrammatical sentences in Figure \ref{fig:ResponseModelUngram}, which show no indication of an interaction between ambiguity and the presence of a plural attractor (($\hat{\beta}=0.06;$ $CI=[-0.25; 0.37];$ $P(\beta<0)=    .36$). It also showed a main effect of plural attractor ($\hat{\beta}=0.50;$ $CI=[0.32; 0.67];$ $P(\beta<0)< .001$). Taken together, the coefficients indicated a substantial agreement attraction effect regardless of the presence of local ambiguity.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[hbt!]

{\centering \includegraphics[width=\linewidth]{figure/ResponseModel-1}

}

\caption{Estimates and 95\% credible intervals for the probit regression coefficients for the model of responses in our experiment and \citet{LagoEtAl:2019}.}\label{fig:ResponseModel}
\end{figure}

\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}[hbt!]

{\centering \includegraphics[width=\linewidth]{figure/ResponseModelUngram-1}

}

\caption{Estimates and 95\% credible intervals for the probit regression coefficients for the model of responses to \protect{\textbf{ungrammatical sentences}} in our experiment and \citet{LagoEtAl:2019}.}\label{fig:ResponseModelUngram}
\end{figure}

\end{knitrout}

### 2022-02-10 paper/lcn_revision/manuscript.Rnw
lcn-revision

<<AverageResponses, message = FALSE, echo=FALSE,  warning = FALSE, dev='cairo_pdf', fig.height=3, fig.align='center', out.width='\\linewidth', fig.pos='hbt!', fig.cap='The average percentage of acceptable responses according to the experimental conditions in our study and \\citet{LagoEtAl:2019}. Error bars signal standard errors calculated following \\citet{Morey:2008,Cousineau:2005}.'>>=

@@

<<ResponseModel, message = FALSE, echo=FALSE,  warning = FALSE, dev='cairo_pdf', fig.height=3, fig.align='center', out.width='\\linewidth', fig.pos='hbt!', fig.cap='Estimates and 95\\% credible intervals for the probit regression coefficients for the model of responses in our experiment and \\citet{LagoEtAl:2019}.'>>=

@@

<<ResponseModelUngram, message = FALSE, echo=FALSE, warning = FALSE, dev='cairo_pdf', fig.height=1.5, fig.align='center', out.width='\\linewidth', fig.pos='hbt!', fig.cap='Estimates and 95\\% credible intervals for the probit regression coefficients for the model of responses to \\protect{\\textbf{ungrammatical sentences}} in our experiment and \\citet{LagoEtAl:2019}.'>>=

>>>>

<<AverageResponses, message = FALSE, echo=FALSE,  warning = FALSE, dev='cairo_pdf', fig.height=3, fig.align='center', out.width='\linewidth', fig.pos='hbt!', fig.cap='The average percentage of acceptable responses according to the experimental conditions in our study and \citet{LagoEtAl:2019}. Error bars signal standard errors calculated following \citet{Morey:2008,Cousineau:2005}.'>>=

@@

<<ResponseModel, message = FALSE, echo=FALSE,  warning = FALSE, dev='cairo_pdf', fig.height=3, fig.align='center', out.width='\linewidth', fig.pos='hbt!', fig.cap='Estimates and 95\% credible intervals for the probit regression coefficients for the model of responses in our experiment and \citet{LagoEtAl:2019}.'>>=

@@

<<ResponseModelUngram, message = FALSE, echo=FALSE, warning = FALSE, dev='cairo_pdf', fig.height=1.5, fig.align='center', out.width='\linewidth', fig.pos='hbt!', fig.cap='Estimates and 95\% credible intervals for the probit regression coefficients for the model of responses to \protect{\textbf{ungrammatical sentences}} in our experiment and \citet{LagoEtAl:2019}.'>>=
